---
title: "Penguin Scatterplot"
format: html
---

## Activity 1: Fill in the Blanks

### Goal

In this activity, we are building a scatterplot that shows the relationship between penguin bill length and penguin bill depth.

We will also color the points by species so we can compare the different penguin groups.

### Step 1: Load the packages

```{r}
library(tidyverse)
library(palmerpenguins)
```

The `tidyverse` package gives us `ggplot2`, which we use to make plots.

The `palmerpenguins` package gives us the `penguins` dataset.

### Step 2: Start the plot

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species
  )
)
```

This starts the plot.

Here is what each part means:

`data = penguins` means we are using the penguins dataset.

`x = bill_length_mm` means bill length goes on the x-axis.

`y = bill_depth_mm` means bill depth goes on the y-axis.

`color = species` means each penguin species gets a different color.

### Step 3: Add the points

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species
  )
) +
  geom_point()
```

`geom_point()` creates a scatterplot.

Each dot is one penguin.

### Step 4: Add labels

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species
  )
) +
  geom_point() +
  labs(
    title = "Bill length and bill depth by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  )
```

The `labs()` function lets us change the plot labels.

The variable name is `bill_depth_mm`, but the reader-friendly label is `"Bill depth (mm)"`.

### Step 5: Add a clean theme

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species
  )
) +
  geom_point() +
  labs(
    title = "Bill length and bill depth by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  theme_minimal()
```

`theme_minimal()` makes the plot look cleaner.

### Step 6: Center the title

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species
  )
) +
  geom_point() +
  labs(
    title = "Bill length and bill depth by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

`hjust = 0.5` centers the title.

Think of it like this:

`hjust = 0` means left aligned.

`hjust = 0.5` means centered.

`hjust = 1` means right aligned.

### Step 7: Add shape for accessibility

This is the bonus part.

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species,
    shape = species
  )
) +
  geom_point() +
  labs(
    title = "Bill length and bill depth by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species",
    shape = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

Now species is shown in two ways:

Color shows species.

Shape also shows species.

This is helpful because some people may have trouble seeing color differences.

### Final Code

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species,
    shape = species
  )
) +
  geom_point() +
  labs(
    title = "Bill length and bill depth by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species",
    shape = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

### What this plot shows

This plot lets us compare bill length and bill depth across penguin species.

Each dot is one penguin.

The x-axis shows bill length.

The y-axis shows bill depth.

The color and shape show the species.

This makes it easier to see whether different penguin species have different bill sizes.

## Activity 1: Fill in the Blanks

Before looking at the completed solution, see if you can fill in each blank yourself.

```{r}
ggplot(
  ___ = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = ___,
    color = ___
  )
) +
  ___() +
  labs(
    title = "Bill length and bill depth by species",
    x = "Bill length (mm)",
    y = "___",
    color = "___"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(___ = 0.5),
    plot.subtitle = element_text(___ = 0.5)
  )
```

### My Answers

| Blank | Your Answer | Correct Answer | Why? |
|:------|:------------|:---------------|:------|
| 1 | | `data` | Tells `ggplot()` which dataset to use. |
| 2 | | `bill_depth_mm` | Places bill depth on the y-axis. |
| 3 | | `species` | Colors each penguin species differently. |
| 4 | | `geom_point()` | Creates a scatterplot. |
| 5 | | `"Bill depth (mm)"` | Human-readable y-axis label. |
| 6 | | `"Species"` | Human-readable legend title. |
| 7 | | `hjust` | Centers the plot title (`0.5` = center). |
| 8 | | `hjust` | Centers the plot subtitle. |

### Check Yourself

Your completed code should look like this:

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species
  )
) +
  geom_point() +
  labs(
    title = "Bill length and bill depth by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```