---
title: "R-LAB 2: Importing Data, Databases, and Web Scraping"
author: "Kameron Suire"
format: html
toc: true
---
# Importing Data into R

## Exercise 1: Importing a CSV

### Why this matters

When importing a CSV, R guesses the data type for every column. Most of the time those guesses are correct, but sometimes you need to override them.

For example, a variable like `year` may be treated as a number even though it is really just a label.

## Load packages

```{r}
library(tidyverse)
```

## Import the data

```{r}
penguins_url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"

penguins <- read_csv(penguins_url)

glimpse(penguins)
```

### What happened?

- `read_csv()` downloads the file from the internet.
- R guesses each column type.
- `glimpse()` lets us quickly inspect the imported dataset.

---

## Force a column type

```{r}
penguins_year_chr <- read_csv(
  penguins_url,
  col_types = cols(
    year = col_character()
  )
)

glimpse(penguins_year_chr)
```

### What happened?

Instead of allowing R to guess, we explicitly told it that `year` should be stored as text.

This is useful whenever a variable is really an identifier rather than a quantity.

---

# Working with Databases

## Why this matters

Large datasets often live inside databases.

Instead of loading millions of rows into R, we send commands to the database and only download the final result.

## Load packages

```{r}
library(DBI)
library(RSQLite)
library(dbplyr)
library(palmerpenguins)
```

## Create an example database

```{r}
con <- dbConnect(SQLite(), ":memory:")

copy_to(
  con,
  palmerpenguins::penguins,
  "penguins",
  temporary = FALSE,
  overwrite = TRUE
)
```

## Query the database

```{r}
penguins_db <- tbl(con, "penguins")

gentoo_summary <- penguins_db |>
  filter(species == "Gentoo") |>
  select(species, island, body_mass_g, sex) |>
  group_by(sex) |>
  summarise(
    mean_body_mass_g = mean(body_mass_g, na.rm = TRUE)
  ) |>
  collect()

gentoo_summary
```

### What happened?

- `tbl()` references the database table.
- `filter()` keeps only Gentoo penguins.
- `select()` keeps only needed columns.
- `group_by()` creates groups.
- `summarise()` calculates averages.
- `collect()` finally downloads the results into R.

The important idea is that **everything before `collect()` happens inside the database.**

---

# Web Scraping

## Why this matters

Unlike a CSV, a webpage is not already a data frame.

Instead, it contains HTML elements such as headings, paragraphs, images, and tables.

We first locate the tables, then convert them into data frames.

## Load packages

```{r}
library(rvest)
library(janitor)
```

## Read a webpage

```{r}
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_population_(United_Nations)"

tables <- url |>
  read_html() |>
  html_elements("table") |>
  html_table()
```

## Inspect the tables

```{r}
length(tables)

glimpse(tables[[1]])
```

### What happened?

- `read_html()` downloads the webpage.
- `html_elements("table")` finds every HTML table.
- `html_table()` converts each table into a data frame.
- `length()` tells us how many tables were found.

Because multiple tables can exist, R stores them inside a **list**.

---

## Clean the first table

```{r}
population_table <- tables[[1]] |>
  select(1:4) |>
  clean_names()

population_table
```

### What happened?

- `tables[[1]]` extracts the first table.
- `select()` keeps the first four columns.
- `clean_names()` creates R-friendly column names.

---

# Key Takeaways

- Always inspect imported data with `glimpse()`.
- Override column types when necessary.
- Use databases for very large datasets.
- Delay `collect()` until the result is small.
- Webpages are hierarchical documents.
- Web scraping often returns lists before data frames.