---
title: "Transform Practice with nycflights13"
format: html
author: "Kameron Suire"
---

## Load Packages

First, we load the packages we need.

```{r}
library(tidyverse)
library(nycflights13)
library(lubridate)
library(gtsummary)
library(flextable)
```

## Starting Data

The `nycflights13` package has several data sets. For this assignment, we start with the `flights` data.

```{r}
flights <- nycflights13::flights
```

## Filter by Air Time

We only want flights that were at least 2 hours but less than 4 hours.

Since `air_time` is measured in minutes:

- 2 hours = 120 minutes
- 4 hours = 240 minutes

```{r}
flights <- flights |>
  filter(
    air_time >= 120,
    air_time < 240
  )
```

## Create Date and Time Variables

The data has separate columns for `year`, `month`, and `day`.

We combine those into one true date variable.

Then we use `dep_time` and `arr_time` to create full departure and arrival timestamps.

The `sprintf("%04d", x)` part makes sure times always have 4 digits.

For example:

- `517` becomes `"0517"`
- `50` becomes `"0050"`

```{r}
flights <- flights |>
  mutate(
    date = make_date(year, month, day),
    
    dep_time_padded = sprintf("%04d", dep_time),
    arr_time_padded = sprintf("%04d", arr_time),
    
    dep_timestamp = ymd_hm(
      paste(
        date,
        str_sub(dep_time_padded, 1, 2),
        str_sub(dep_time_padded, 3, 4)
      )
    ),
    
    arr_timestamp = ymd_hm(
      paste(
        date,
        str_sub(arr_time_padded, 1, 2),
        str_sub(arr_time_padded, 3, 4)
      )
    )
  )
```

## Filter by Flight Details

Now we keep only flights going to:

- Orlando = `MCO`
- Fort Lauderdale = `FLL`
- Miami = `MIA`

And only flights from:

- American Airlines = `AA`
- Delta Air Lines = `DL`
- United Airlines = `UA`

```{r}
flights <- flights |>
  filter(
    dest %in% c("MCO", "FLL", "MIA"),
    carrier %in% c("AA", "DL", "UA")
  )
```

## Create Factor Variables

Right now, `carrier` and `dest` are short codes.

We turn them into factor variables with easier-to-read labels.

```{r}
flights <- flights |>
  mutate(
    carrier = factor(
      carrier,
      levels = c("AA", "DL", "UA"),
      labels = c("American Airlines", "Delta Air Lines", "United Airlines")
    ),
    
    dest = factor(
      dest,
      levels = c("MCO", "FLL", "MIA"),
      labels = c("Orlando", "Fort Lauderdale", "Miami")
    )
  )
```

## Join Flights with Weather

The weather data has temperature and wind speed.

To join weather to flights, we match using:

- `origin`
- `time_hour`

These tell R which airport and hour the weather belongs to.

```{r}
weather_small <- nycflights13::weather |>
  select(origin, time_hour, temp, wind_speed)

flights <- flights |>
  left_join(weather_small, by = c("origin", "time_hour"))
```

## Check the Final Data

This lets us confirm that `temp` and `wind_speed` were added.

```{r}
names(flights)
```

## Summary Table

Finally, we make a table that summarizes destination, temperature, wind speed, time in the air, and distance by airline.

```{r}
flights |>
  dplyr::select(carrier, dest, temp, wind_speed, air_time, distance) |>
  gtsummary::tbl_summary(
    by = carrier,
    missing = "no",
    statistic = gtsummary::all_continuous() ~ "{mean} ± {sd}",
    type = distance ~ "continuous",
    label = list(
      dest ~ "Destination",
      temp ~ "Temperature",
      wind_speed ~ "Wind Speed",
      air_time ~ "Time in the Air",
      distance ~ "Distance"
    )
  ) |>
  gtsummary::as_flex_table()
```

## What This Assignment Practiced

In this assignment, we practiced several important `dplyr` skills:

- `filter()` to keep certain rows
- `mutate()` to create new variables
- `factor()` to make readable category labels
- `select()` to keep certain columns
- `left_join()` to combine two data sets
- `tbl_summary()` to make a clean summary table